---
title: "TidyModels - PCA and UMAP"
description: |
  Working with TidyModels to generate PCA and UMAP for Cocktail recipes. Using these
  recipes extrapolating the relation of a cocktail to each other.
author:
  - name: Karat Sidhu
    url: {}
date: 2022-05-22
preview: images/logo.svg
categories:
  - TidyModels
  - PCA
  - UMAP
  - Machine Learning
  - Dimension Reduction
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# TidyModels - PCA & UMAP

## Dimensionality Reduction analysis with Tidymodels

**Note: This is part 2 of the tidymodels practise. A few of these RMarkdown files were generated on a date different to the publishing date.**

Principal Component analysis (and UMAP to a lesser extent) have been extensively used in Biological Sciences 
and specifically in Omics. Compared to other components of the TidyVerse, I am fairly comfortable using and running the PCA test
on a set of data. However, I typically use either the base R (`prcom`) or PCA Tools^[https://bioconductor.org/packages/release/bioc/html/PCAtools.html] because generating a series of plots using those packages
is fairly straightforward^[https://github.com/SidhuK/R_for_Metabolomics/tree/main/Principal_Component_Analysis].



# Cocktail Recipes Dataset

The dataset used is from the Tidytuesday data, where different cocktails and the components used for their recipes is provided.  
Reducing dimensions in such a case would help in finding out which cocktail is closely related to the other in terms of their components.


# Loading packages

```{r}
library(tidyverse)
library(tidymodels)
library(tidytext)

```



# Loading Data

```{r}
boston_cocktails <-
  readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-26/boston_cocktails.csv")
```



# Data Cleaning and Exploratory Data Analysis

Preliminary look at the data

```{r}

boston_cocktails
```

So the PCA should be carried out for the "name" field and looking at the relation between the cocktail and the incredients, category and the measured amount of each ingredient used. 

## Most Common Ingredients

```{r}

boston_cocktails |>
  count(ingredient, sort = TRUE)
```

A lot of cocktails look like are using Gin, Lemon Juice, Syrup and Vodka as some of the most common ingredients. 


## Column Cleaning and data conversion

The data isn't very clean yet. Some ingredients, which are clearly the same are labelled slightly differently. 
So we need to clean and make the data usable for our analysis. 

Some of the main changes needed in the **ingredients**:

- turn all ingredients to lowercase

- "-" needs to be chagned to " "

- "lemon" to "lemon Juice"

- "lime" to "lime juice"

- "grapefruit" to "grapefruit juice" (same for orange)


Some of the main changes needed in the **measure**:

- 1/2 to 0.5

- 3/4 to 0.75

- 1/4 to 0.25

- remove words like "dash", and "oz" etc.

- Convert the entire column into a number



```{r}

cocktails <- boston_cocktails |> 
  mutate(
    ingredient = str_to_lower(ingredient),
    ingredient = str_replace_all(ingredient, "-", " "),
    ingredient = str_remove(ingredient, " liqueur$"),
    ingredient = str_remove(ingredient, " (if desired)$"),
    ingredient = case_when(
      str_detect(ingredient, "bitters") ~ "bitters",
      str_detect(ingredient, "lemon") ~ "lemon juice",
      str_detect(ingredient, "lime") ~ "lime juice",
      str_detect(ingredient, "grapefruit") ~ "grapefruit juice",
      str_detect(ingredient, "orange") ~ "orange juice",
      TRUE ~ ingredient
    ),
    measure = case_when(
      str_detect(ingredient, "bitters") ~ str_replace(measure, "oz$", "dash"),
      TRUE ~ measure
    ),
    measure = str_replace(measure, " ?1/2", ".5"),
    measure = str_replace(measure, " ?3/4", ".75"),
    measure = str_replace(measure, " ?1/4", ".25"),
    measure_number = parse_number(measure),
    measure_number = if_else(str_detect(measure, "dash$"), # a few drops = so 0.02 oz
      measure_number / 50,
      measure_number
    )
  ) |> 
  add_count(ingredient) |> 
  filter(n > 15) |> 
  select(-n) |> 
  distinct(row_id, ingredient, .keep_all = TRUE) |> 
  na.omit()


cocktails

```

## New Dataframe with a "wide" format

We're now close to beginning our analysis. However, the data is in long format 
and PCA prefers to get the data wider. So we'll pivot_wide to make the analysis easier.

```{r}

cocktails_df <- cocktails |> 
  select(-ingredient_number, -row_id, -measure) |> 
  pivot_wider(names_from = ingredient, values_from = measure_number, values_fill = 0) |> 
  janitor::clean_names() |> 
  na.omit()

cocktails_df
```

This will be the starting dataset used to run PCA (and UMAP)


# Principal Component Analysis

Henceforth I will be using CMDLineTips Post^[https://cmdlinetips.com/2020/06/pca-with-tidymodels-in-r/] as a reference
because I've never done a PCA with TidyModels before.

## Prepare the analysis

### Recipe code

```{r}
pca_rec <- recipe(~., data = cocktails_df) |>  # what data to use
  update_role(name, category, new_role = "id") |> # name and category are identifiers not variables
  step_normalize(all_predictors()) |> # normalize all other columns
  step_pca(all_predictors()) # pca for all other columns


```

### Prep Code

```{r}
pca_prep <- prep(pca_rec)

pca_prep


```

The prep object now doesn't contain any output, and we need to tidy the object to read it.
List object contains the PCA "results" and components on the 2nd place under "term_info" (open it in the console to see details).

## PCA Table

```{r}
tidied_pca <- tidy(pca_prep, 2)

tidied_pca

```
## PCA Plots

A lot of Principal components seen, plotting the first 4

```{r fig.height=8}

tidied_pca |> 
  filter(
    component == "PC1" |
      component == "PC2" |
      component == "PC3" |
      component == "PC4"
   ) |> 
  mutate(component = fct_inorder(component)) |> 
    ggplot(aes(value, terms, fill = terms)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~component, nrow = 1) +
  hrbrthemes::theme_ipsum() +
  labs(y = NULL) 

```

Looking at the PC1, simple syrup and powdered sugar are very different from each other, therefore the cocktails 
must be using one or the other. They also appear to be the two biggest factors.

### PC1 and PC2 main contributors

```{r}
tidied_pca |> 
  filter(component %in% paste0("PC", 1:2)) |> 
  group_by(component) |>
  top_n(8, abs(value)) |>
  ungroup() |>
  mutate(terms = reorder_within(terms, abs(value), component)) |>
  ggplot(aes(abs(value), terms, fill = value > 0)) +
  geom_col() +
  facet_wrap(~component, scales = "free_y") +
  scale_y_reordered() +
  labs(
    x = "Absolute value of contribution",
    y = NULL, fill = "Positive?"
  ) +
  hrbrthemes::theme_ipsum()

```


### PCA Plot



```{r}

juice(pca_prep) |> 
  ggplot(aes(PC1, PC2, label = name)) +
  geom_point(aes(color = category), alpha = 0.7, size = 2) +
  ggrepel::geom_text_repel(max.overlaps = 10) +
  labs(color = NULL) + hrbrthemes::theme_ipsum()
  

```

Similar kind of cocktails are clustered together, so PCA analysis appears to be somewhat successful.


# Uniform Manifold Automation & Projection

## Recipe Prep

The embed package provides recipe steps for ways to create embeddings including UMAP.

```{r}
library(embed)

umap_rec <- recipe(~., data = cocktails_df) |>
  update_role(name, category, new_role = "id") |>
  step_normalize(all_predictors()) |>
  step_umap(all_predictors())

umap_prep <- prep(umap_rec)

umap_prep

```

So far, the process is exactly the same as PCA. Prep the recipe and get the "non output"-output.

```{r}

juice(umap_prep) |> 
  ggplot(aes(UMAP1, UMAP2, label = name)) +
  geom_point(aes(color = category), alpha = 0.7, size = 2) +
  ggrepel::geom_text_repel(max.overlaps = 30) +
  labs(color = NULL) + hrbrthemes::theme_ipsum()
 

```


Similarly clustering seen here, but type of cocktails seen here are different. 

# Conclusion

The PCA (and UMAP) using TidyModels offers a bit more customization compared to using a regular 
`prcomp` or `PCA-Tools`.  However, it requires a bit more code and felt slightly slower in running 
compared to the other packages (without actually testing the time fyi)^[Plots were slower to render].



# Additional Reading:

- Easily the best explanation of PCA on the internet:
https://stats.stackexchange.com/questions/2691/making-sense-of-principal-component-analysis-eigenvectors-eigenvalues

- How PCA works - interactive plots:
https://setosa.io/ev/

- How UMAP works:
https://umap-learn.readthedocs.io/en/latest/how_umap_works.html




