---
title: "EPL 2022 Final Table"
description: |
  Using gt package to create a great looking table for the EPL Standings 2021-22.
author:
  - name: Karat Sidhu
    url: {}
date: 2022-05-23
preview: images/logo.png
categories:
  - Data-Viz
  - misc
  - TidyVerse
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# How to Make beautiful tables in R using gt() 

## Using gt package to create a great looking table for the EPL Standings 2021-22


Grammar of tables or gt() package, combined with the gtextras() package provides an easy and simple way to make high quality tables in R.

I wanted to  practice the gt() package in R, and decided to document my journey to making a great looking table of the final standings for the English Premier League now that the season is over.


It’s been a season to forget as a Manchester United fan :(   and I am hopeful for better performances next season under Ten Hag, but the football as a neutral has been very exciting to watch.I digress, let’s get back on to the coding part.


Code to make this possible, along with a brief explanation of each part of the process:


# Loading the required packages

```{r}
library(tidyverse)
library(gt)
library(gtExtras)
library(showtext)
library(RColorBrewer)


```



The main font I used for the title is similar to the PL official font. 

```{r}

showtext_auto()
sysfonts::font_add("Premier League", regular = "FontsFree-Net-Barclays-Premier-League.ttf")

```




The data file was downloaded from the FBref database as a .csv, so the loading and cleaning of data was a relatively easy process

# Loading the data

```{r}

table <- read.csv("pl_table_gw_38.csv")


logos <-
    read.csv(
        "https://raw.githubusercontent.com/steodose/Club-Soccer-Forecasts/main/team_mapping.csv"
    )



```



Data to be cleaned and the form table needs to be in the form of a list inside a data frame. 
Cleaning it up:

```{r}

logos_cleaned <- logos |>
    select(squad_fbref, url_logo_espn) |>
    rename(Squad = "squad_fbref")

table <- table |>
    select(-c(Goalkeeper, Notes, Attendance))

```





# Combining the two tables, 


```{r}

table2 <- inner_join(table,
    logos_cleaned,
    by = "Squad"
) |>
    relocate(Rk, url_logo_espn)


```



# Making the final table:

```{r}

final_table <- table2 |>
    rename(` ` = "url_logo_espn") |>
    rename(Rank = "Rk") |>
    rename(`Top Scorer` = "Top.Team.Scorer") |>
    gt() |>
    gt_color_rows(
        columns = W,
        type = "discrete",
        palette = "RColorBrewer::Spectral",
        domain = c(max(table2$W), min(table2$W))
    ) |>
    gt_color_rows(
        columns = D,
        type = "discrete",
        direction = -1,
        palette = "RColorBrewer::Spectral",
        domain = c(max(table2$D), min(table2$D))
    ) |>
    gt_color_rows(
        columns = L,
        type = "discrete",
        direction = -1,
        palette = "RColorBrewer::Spectral",
        domain = c(max(table2$L), min(table2$L))
    ) |>
    gt_color_rows(
        columns = GF,
        type = "discrete",
        palette = "RColorBrewer::Spectral",
        domain = c(max(table2$GF), min(table2$GF))
    ) |>
    gt_color_rows(
        columns = GA,
        type = "discrete",
        palette = "RColorBrewer::Spectral",
        direction = -1,
        domain = c(min(table2$GA), max(table2$GA))
    ) |>
    gt_color_rows(
        columns = GD,
        type = "discrete",
        palette = "RColorBrewer::Spectral",
        domain = c(max(table2$GD), min(table2$GD))
    ) |>
    gt_color_rows(
        columns = Pts,
        type = "discrete",
        palette = "RColorBrewer::Spectral",
        domain = c(max(table2$Pts), min(table2$Pts))
    ) |>
    gt_color_rows(
        xG,
        type = "discrete",
        palette = "RColorBrewer::Spectral",
        domain = c(max(table2$xG), min(table2$xG))
    ) |>
    gt_color_rows(
        xGA,
        type = "discrete",
        palette = "RColorBrewer::Spectral",
        direction = -1,
        domain = c(min(table2$xGA), max(table2$xGA))
    ) |>
    gt_color_rows(
        xGD,
        type = "discrete",
        palette = "RColorBrewer::Spectral",
        domain = c(max(table2$xGD), min(table2$xGD))
    ) |>
    gt_color_rows(
        xGD.90,
        type = "discrete",
        palette = "RColorBrewer::Spectral",
        domain = c(max(table2$xGD.90), min(table2$xGD.90))
    ) |>
    gt_img_rows(
        columns = ` `,
        img_source = "web",
        height = 35
    ) |>
    tab_style(
        style = list(cell_text(weight = "bold")),
        locations = cells_body(rows = Squad == "Manchester Utd")
    ) |>
    gt_theme_espn() |>
    tab_style(
        style = list(cell_text(weight = "bold")),
        locations = cells_body(
            columns = Squad
        )
    ) |>
    tab_header(
        title = "2021-22 Premier League Standings",
        subtitle = "Final Results for the English Premier League saw Manchester City win the league after an exciting last few weeks of competition
    with Liverpool. The competition came down to the final few minutes of the season where City were crowned winners.
    The top 4 teams qualified for the UCL, 5th and 6th place teams are in the Europa League while the 7th place West Ham
    secured the UECL qualification. The bottom 3 teams, Norwich, Watford and Burnley were relegated to the English Championship (2nd tier competition);
    Everton and Leeds United narrowly escaped relegation."
    ) |>
    tab_footnote(
        footnote = "xG = Expected Goals (Higher is better)",
        locations = cells_column_labels(columns = xG)
    ) |>
    tab_footnote(
        footnote = "xGA = Expect Goals Against (Lower is better) ",
        locations = cells_column_labels(columns = xGA)
    ) |>
    tab_footnote(
        footnote = "xGD = xG Difference (xG - xGA (High is better))",
        locations = cells_column_labels(columns = xGD)
    ) |>
    tab_footnote(
        footnote = "xGD.90 = xG Difference per 90 mins (Higher is better)",
        locations = cells_column_labels(columns = xGD.90)
    ) |>
    tab_footnote(
        footnote = "The team that I support :(",
        locations = cells_body(
            columns = Squad,
            rows = 6
        )
    ) |>
    tab_source_note(
        source_note = "Data : FBRef | Table : Github.com/SidhuK"
    ) |>
    tab_style(
        locations = cells_title(groups = "title"), # format the main title
        style = list(
            cell_text(
                font = "Premier League",
                size = px(40),
                color = "black",
                weight = 700
            )
        )
    ) |>
    tab_options(
        table.background.color = "#eeeeee",
        column_labels.background.color = "#eeeeee"
    ) # set the bg color


```





# Saving the table as png (optional)


(Use gt save to save it as a png and not an html object)



# Final Result:

```{r}
final_table

```












